Using language as an indicator of cognitive decline in pre-clinical Alzheimer's Disease.

### 2020-04-06 Source/Paper 1 - Presidents Corpus Paper/PresidentsCorpus.tex
updated Pres

\maketitle
\section{Introduction}\label{introduction}
A diagnosis of dementia is generally made when there is a decline in brain function due to physical changes in the brain \cite{Albert2011}. It affects a significant proportion of the global older adult population and the impact on morbidity and mortality rates is considerable. Dementia including Alzheimer's Disease is currently the leading cause of death in England and Wales \cite{Patel2018} and the sixth leading cause of death in the United States (US) accounting for 32\% of all adult deaths in the US and projected to rise to 43\% by 2050 \cite{Weuve2014} . A 2014 report commissioned by the Alzheimer's Society estimated that in the UK by 2015 there would be approximately 855,000 people rising living with dementia increasing to 1 million by 2021 \cite{AlzheimersSociety2014}. This represents 1 in 79 of the total UK population rising to 1 in 14 of those aged 65 or over \cite{AlzheimersSociety2014}. Worldwide, there are 46 million people with a diagnosis of dementia globally and that number is expected to hit 131.5 million by 2050 \cite{Prince2015}. From a financial perspective, the cost burden is also significant. The estimated annual spend on dementia healthcare in the UK is £4.3 billion of which approximately £85 million is spent on diagnosis. The total financial burden of dementia (excluding the costs associated with early onset dementia) is £26.3 billion annually. Globally, this picture is a lot bleaker. The worldwide cost of dementia in 2018 was estimated to be in the region of one trillon US dollars \cite{Prince2015}.
There are different types of dementia including Alzheimer's Disease (AD), vascular dementia, dementia with Lewy bodies and fronto-temporal dementia, all of which are currently incurable. AD is a progressive neurodegenerative disease and is the most common type of dementia, responsible for approximately 60\% to 80\% of all cases \cite{Duong2017}. Currently, a definitive diagnosis for AD can only be produced post-mortem. However, there are a number of psychological and physiological indicators that can indicate that AD is is present. These physiological changes lead to the development of some of the psychological symptoms associated with AD, primarily cognitive deficits such as problems with episodic and semantic memory, organizing and planning and other problems with executive function, difficulties with language, and visuospatial deficits \cite{McKhann2011}. These cognitive symptoms are often accompanied by emotional problems such as depression and behavioural difficulties. As more neurons die throughout the brain, a person with AD gradually loses the ability to think, remember, make decisions and function independently.
Despite the increasing prevalence of AD and an improved understanding about how it affects the brain there are no medications that improve prognosis. All the medications that are currently on the market are designed to manage symptoms. Whilst there are numerous investigational drugs in development for the treatment of AD, a larger than normal percentage of these drugs fail in the clinical trial stage of the drug discovery process (99.6\% failure rate vs 80\% for systemic anti-cancer drugs) \cite{Cummings2014}. Cummings et al proposed that a possible reason for the lack of success is that the drug treatments are initiated too far along in the progression of the disease and thus much of the degeneration of the brain has already occurred \cite{Cummings2014}. Research focus has now started to shift to the earlier stages of AD (i.e. symptomatic pre-dementia phase of AD) which some literature describes as 'Mild Cognitive Impairment (MCI) due to AD'.
One of the challenges associated with the early detection of AD is differentiating natural age associated memory impairment and cognitive decline due to aging from decline due to AD \cite{Lo2017}. This challenge is often complicated further due to the large variation in the cognitive abilities and educational background of individuals \cite{Harada2013}. The work of Albert et al helps to address this by the development of clinical criteria which professionals can use to diagnose MCI due to AD. One of the most important observations from this piece of work is that a diagnosis of MCI requires evidence of intra-individual change and optimally requires evaluation at two or more points \cite{Albert2011}.This is essentially to place more importance on the trajectory of a person's cognitive abilities rather than a person's cognitive ability in general. Berisha and Liss used the transcripts of press conferences of Ronald Reagan who would go on to be diagnosed with AD, and George Bush Sr who had no diagnosis. These speeches are in the public domain and provide a good opportunity of analysing the spontaneous language use of someone with AD, before a diagnosis was made. 
There has been significant research in the area of language deterioration as a method of detecting AD at an earlier stage. This usually takes the form of recording speech whilst patients' undertake a cognitive assessment such as the Picture Description Task \cite{Fraser2015}. Given that language samples are relatively easy to collect, research has moved towards analysis of spontaneous speech. The work of Berisha and Liss is a good example of this, their study compared the differences in language use between two US presidents, Ronald Reagan (RR), who would go on to receive a diagnosis of Dementia and George H. W. Bush (GHWB) who acted as a matched control based on Age \cite{Berisha2015}. They found several differences in language use over time which they felt acted as indicators of RR's difficulties with language due to AD. Differences in features of language identified to be statistically significant included the number of unique words used per speech, the use of non-specific nouns and fillers and low imageability verbs \cite{Berisha2015}. 
Berisha and Liss have developed interesting ideas about how we might track changes over time in various lexical features which have been associated with the development of Alzheimer's Disease. However we feel that cognitive decline in pre-clinical AD is not accurately modeled as a linear process and therefore we explored the application of generalised additive models (GAMs) to this data which have no assumed understanding of the distribution of the data. We refine these ideas with the aim of exploring the potential for a protocol that can be used for analyzing language deterioration.

\begin{figure}[H]
	\centering
	\includegraphics[width=6cm, height=5cm]{plots/comparison1.png}
	\caption{Comparing linear model with a non-linear model}
\end{figure}

@@

\subsection{Comparing linear models and non-linear models}
In terms of comparing the general additive model and linear models. We calculated the predicted residual error sum of squares (PRESS) statistic for each feature. The PRESS statistic
is a form of cross-validation used in regression analysis to provide a summary measure of the fit of a model to a sample of observations that were not themselves used to estimate the model. There were no significant differences when comparing the models (paired t-test, t = 1.8875, df = 21, p = 0.07299).

@@

\begin{tabular}{lrr}

@@

feature & gamPRESS & lmPRESS \\ 

@@

ppron & 18.59 & 20.00 \\ 
  social & 21.68 & 24.91 \\ 
  NounsNormalised & 0.00 & 0.00 \\ 
  function. & 46.55 & 46.03 \\ 
  conj & 8.81 & 8.35 \\ 
  PronounsNormalised & 0.00 & 0.00 \\ 
  Analytic & 2272.20 & 2239.22 \\ 
  NN & 101685.87 & 96485.58 \\ 
  male & 5.19 & 5.28 \\ 
  UniqueWords & 255114.64 & 238042.43 \\ 
  WDT & 1987.15 & 1958.78 \\ 
  Nouns/100 & 27.67 & 26.22 \\ 
  UniqueStems & 170450.65 & 166302.73 \\ 
  shehe & 5.19 & 5.27 \\ 
  VBZ & 13889.87 & 13721.18 \\ 
  JJ & 24950.44 & 24692.04 \\ 
  article & 13.50 & 13.47 \\ 
  Adjectives.100 & 3.02 & 2.99 \\ 
  Dic & 40.10 & 38.50 \\ 

@@

Another we did was to look at the difference in decline between the presidents. We compared both RR v GHWB and RR v DJT. We found that for those features that were significant in terms of language decline for RR, they were significantly different for both our controls. This does appear to mean that for RR, his language changed in a considerably different way as he got older and we could attribute this decline to a pre-clinical stage of alzheimer's disease.

@@

We can clearly see that a linear model is insufficient for modelling this particular case and whilst Iris Murdoch may be an atypical case, we can see the value in using non-linear models that may be able to map these unusual cases whilst retaining the ability to illustrate fluctuations in language use.

@@

In her paper on developing diagnositic criteria for Mild Cognitive Impairment, Albert et all \cite{Albert2011} states 'it is important to obtain longitudinal assessments of cognition, whenever possible' and 'obtaining objective evidence of progressive declines in cognition over time is important for establishing the accuracy of the diagnosis, as well as for

>>>>

\maketitle
\section{Introduction}\label{introduction}
A diagnosis of dementia is generally made when there is a decline in brain function due to physical changes in the brain \cite{Albert2011}. It affects a significant proportion of the global older adult population and the impact on morbidity and mortality rates is considerable. Dementia including Alzheimer's Disease is currently the leading cause of death in England and Wales \cite{Patel2018} and the sixth leading cause of death in the United States (US) accounting for 32\% of all adult deaths in the US and projected to rise to 43\% by 2050 \cite{Weuve2014} . A 2014 report commissioned by the Alzheimer's Society estimated that in the UK by 2015 there would be approximately 855,000 people rising living with dementia increasing to 1 million by 2021 \cite{AlzheimersSociety2014}. This represents 1 in 79 of the total UK population rising to 1 in 14 of those aged 65 or over \cite{AlzheimersSociety2014}. Worldwide, there are 46 million people with a diagnosis of dementia globally and that number is expected to hit 131.5 million by 2050 \cite{Prince2015}. From a financial perspective, the cost burden is also significant. The estimated annual spend on dementia healthcare in the UK is £4.3 billion of which approximately £85 million is spent on diagnosis. The total financial burden of dementia (excluding the costs associated with early onset dementia) is £26.3 billion annually. Globally, this picture is a lot bleaker. The worldwide cost of dementia in 2018 was estimated to be in the region of one trillon US dollars \cite{Prince2015}.
\par 
There are different types of dementia including Alzheimer's Disease (AD), vascular dementia, dementia with Lewy bodies and fronto-temporal dementia, all of which are currently incurable. AD is a progressive neurodegenerative disease and is the most common type of dementia, responsible for approximately 60\% to 80\% of all cases \cite{Duong2017}. Currently, a definitive diagnosis for AD can only be produced post-mortem. However, there are a number of psychological and physiological indicators that can indicate that AD is is present. These physiological changes lead to the development of some of the psychological symptoms associated with AD, primarily cognitive deficits such as problems with episodic and semantic memory, organizing and planning and other problems with executive function, difficulties with language, and visuospatial deficits \cite{McKhann2011}. These cognitive symptoms are often accompanied by emotional problems such as depression and behavioural difficulties. As more neurons die throughout the brain, a person with AD gradually loses the ability to think, remember, make decisions and function independently.
\par
Despite the increasing prevalence of AD and an improved understanding about how it affects the brain there are no medications that improve prognosis. All the medications that are currently on the market are designed to manage symptoms. Whilst there are numerous investigational drugs in development for the treatment of AD, a larger than normal percentage of these drugs fail in the clinical trial stage of the drug discovery process (99.6\% failure rate vs 80\% for systemic anti-cancer drugs) \cite{Cummings2014}. Cummings et al proposed that a possible reason for the lack of success is that the drug treatments are initiated too far along in the progression of the disease and thus much of the degeneration of the brain has already occurred \cite{Cummings2014}. 
\par

Research focus has now started to shift to the earlier stages of AD (i.e. symptomatic pre-dementia phase of AD) which some literature describes as 'Mild Cognitive Impairment (MCI) due to AD'.
\par
One of the challenges associated with the early detection of AD is differentiating natural age associated memory impairment and cognitive decline due to aging from decline due to AD \cite{Lo2017}. This challenge is often complicated further due to the large variation in the cognitive abilities and educational background of individuals \cite{Harada2013}. The work of Albert et al helps to address this by the development of clinical criteria which professionals can use to diagnose MCI due to AD. One of the most important observations from this piece of work is that a diagnosis of MCI requires evidence of intra-individual change and optimally requires evaluation at two or more points \cite{Albert2011}.This is essentially to place more importance on the trajectory of a person's cognitive abilities rather than a person's cognitive ability in general. Berisha and Liss used the transcripts of press conferences of Ronald Reagan who would go on to be diagnosed with AD, and George Bush Sr who had no diagnosis. These speeches are in the public domain and provide a good opportunity of analysing the spontaneous language use of someone with AD, before a diagnosis was made. 
\par 
There has been significant research in the area of language deterioration as a method of detecting AD at an earlier stage. This usually takes the form of recording speech whilst patients' undertake a cognitive assessment such as the Picture Description Task \cite{Fraser2015}. Given that language samples are relatively easy to collect, research has moved towards analysis of spontaneous speech. The work of Berisha and Liss is a good example of this, their study compared the differences in language use between two US presidents, Ronald Reagan (RR), who would go on to receive a diagnosis of Dementia and George H. W. Bush (GHWB) who acted as a matched control based on Age \cite{Berisha2015}. They found several differences in language use over time which they felt acted as indicators of RR's difficulties with language due to AD. Differences in features of language identified to be statistically significant included the number of unique words used per speech, the use of non-specific nouns and fillers and low imageability verbs \cite{Berisha2015}. 
\par
Berisha and Liss have developed interesting ideas about how we might track changes over time in various lexical features which have been associated with the development of Alzheimer's Disease. However we feel that cognitive decline in pre-clinical AD is not accurately modeled as a linear process and therefore we explored the application of generalised additive models (GAMs) to this data which have no assumed understanding of the distribution of the data. We refine these ideas with the aim of exploring the potential for a protocol that can be used for analyzing language deterioration.

\begin{figure}[H]
	\centering
	\includegraphics[width=12cm, height=9cm]{plots/comparison1.png}
	\caption{Comparing linear model with a non-linear model}
\end{figure}

@@

\subsection{Comparing linear models and non-linear models}
In terms of comparing the general additive model and linear models. We calculated the predicted residual error sum of squares (PRESS) statistic for each feature. The PRESS statistic is a form of leave one out cross-validation (LOOCV) used in regression analysis to provide a summary measure of the fit of a model to a sample of observations that were not themselves used to estimate the model. There were no significant differences when comparing the models (paired t-test, t = 1.8875, df = 21, p = 0.07299).

@@

\begin{tabular}{| l | r | r | r |}

@@

feature & gamPRESS & lmPRESS & Best Model\\ 

@@

ppron & 18.59 & 20.00 & GAM\\ 
  social & 21.68 & 24.91 & GAM\\ 
  NounsNormalised & 0.00281 & 0.00279 & LM\\ 
  function. & 46.55 & 46.03 & LM\\ 
  conj & 8.81 & 8.35 & LM \\ 
  PronounsNormalised & 0.00228 & 0.00235 & GAM \\ 
  Analytic & 2272.20 & 2239.22 & LM\\ 
  NN & 101685.87 & 96485.58 & LM\\ 
  male & 5.19 & 5.28 & GAM\\ 
  UniqueWords & 255114.64 & 238042.43 & LM\\ 
  WDT & 1987.15 & 1958.78 & LM \\ 
  Nouns/100 & 27.67 & 26.22 & LM\\ 
  UniqueStems & 170450.65 & 166302.73 & LM \\ 
  shehe & 5.19 & 5.27 & GAM\\ 
  VBZ & 13889.87 & 13721.18 & LM\\ 
  JJ & 24950.44 & 24692.04 & LM\\ 
  article & 13.50 & 13.47 & LM\\ 
  Adjectives.100 & 3.02 & 2.99 & LM\\ 
  Dic & 40.10 & 38.50 & LM\\ 

@@

Another we did was to look at the difference in decline between the presidents. We compared both RR v GHWB and RR v DJT. We found that for those features that were significant in terms of language decline for RR, they were significantly different for both our controls. This does appear to mean that for RR, his language changed in a considerably different way as he got older and we could attribute this decline to a pre-clinical stage of AD.

@@

We can clearly see that a linear model is insufficient for modeling this particular case and whilst Iris Murdoch may be an atypical case, we can see the value in using non-linear models that may be able to map these unusual cases whilst retaining the ability to illustrate fluctuations in language use.

@@

In her paper on developing diagnostic criteria for Mild Cognitive Impairment, Albert et al \cite{Albert2011} states 'it is important to obtain longitudinal assessments of cognition, whenever possible' and 'obtaining objective evidence of progressive declines in cognition over time is important for establishing the accuracy of the diagnosis, as well as for
